\documentclass[11pt]{formatting-template}

% mathptmx is a Times Roman look-alike (don't use the times package)
% It isn't clear if Times is required. The OGS manual lists several
% "standard fonts" but never says they need to be used.
% \usepackage{mathptmx}

\usepackage[NoDate]{currvita}
\usepackage{array}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{ragged2e}
\usepackage{microtype}
\usepackage[breaklinks=true,pdfborder={0 0 0}]{hyperref}
\usepackage{graphicx}
\AtBeginDocument{%
	\settowidth\cvlabelwidth{\cvlabelfont 0000--0000}
}

% OGS recommends increasing the margins slightly.
\increasemargins{.1in}

\usepackage[english]{babel}
\usepackage{blindtext}

\usepackage{url}

\usepackage[super,sort&compress]{natbib}

\overfullrule5pt

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Thesis Title & Author Information
\title{Decoding the genomic regulatory syntax driving notochord development}
\author{Michelle Franc Ragsac}
\degree{Bioinformatics \& Systems Biology}{Doctor of Philosophy}

% Thesis Committee Members
\chair{Professor Emma K. Farley}
\cochair{Professor Theresa Gaasterland} 
\committee{Professor Vineet Bafna}
\committee{Professor Christopher Benner} % alphabetical order required
\committee{Professor Xin Sun}
\degreeyear{2022}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

% Begin with front matter and so forth
\frontmatter
\maketitle{}
\makecopyright{}
\makesignature{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dedication
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{dedication}
	\setsinglespacing{}
	\parindent0pt\parskip\baselineskip{}
	\vskip0pt plus.25fil
	\begin{center}
		\textit{\textbf{Ang pamilya ay hindi isang mahalagang bagay lamang.\\ Ito ay ang lahat.}}

		$\cdots$

		This dissertation is dedicated to the family that fostered a warm,\\
		multi-generational home that valued love, curiosity, and education above all else.
		
		To my parents, \textit{Francisco Gregorio Ragsac} and \textit{Lea Reyes Ragsac},\\
		for paving the way for a better life in the United States.
		
		To my younger brother, \textit{Thomas Jonathan Ragsac},\\
		for spending time with me through playing video games because you had to. 

		To my aunt, \textit{Charleen Rodriguez Ragsac},\\
		for always keeping things loud and lively.
		
		Finally, to my grandparents, \textit{Hermogenes Riego de Dios Reyes} and \textit{Josefina Tardeo Reyes},\\
		for making sure I spent my days listening to classical music on the mandolin and eating comforting Filipino food during my youth.

		$\cdots$

		I would also like to dedicate this dissertation to \textit{Clarence Kuang-Le Mah},\\
		the nerdy best friend who got me interested in bioinformatics in the first place.\\
		I’d still walk from Camp Snoopy to the Village for you.
	\end{center}
\end{dedication}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Epigraph
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{epigraph}
	\vskip0pt plus.5fil
	\vfil\vfil
	\setsinglespacing{}
	{ \flushright{
		It is not birth, marriage, or death, \\
		but gastrulation which is the most \\
		important time in your life.
		}
		\vskip\baselineskip{}
		\textit{Louis Wolpert}\par
	}
	\vfil\vfil\vfil
\end{epigraph}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Table of Contents, Figures, Tables, Etc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Next comes the table of contents, list of figures, list of tables,
% etc. If you have code listings, you can use \listoflistings (or
% \lstlistoflistings) to have it be produced here as well. Same with
% \listofalgorithms.

\tableofcontents
\listoffigures
\listoftables

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{acknowledgements}
	First and foremost, I have to thank my thesis advisor and committee chair, \textit{Professor Emma Farley}. Emma is one of the brightest scientists I’ve met, and my approach to research has definitely been shaped by her guidance over the past few years. Her endless support throughout graduate school-—especially during difficult moments in research and my personal life—ultimately helped me get to where I am today. I would also like to thank my long-time advisor and thesis committee co-chair, \textit{Professor Theresa “Terry” Gaasterland}. I have known Terry since my early undergraduate years at UCSD, and her infectious enthusiasm for interesting problems in genetics and genomics is something that I will always carry with me. I would also like to show gratitude to the rest of my committee, \textit{Professors Vineet Bafna}, \textit{Christopher Benner}, and \textit{Xin Sun}. Without their assistance and helpful feedback throughout the later portion of my graduate school journey, this work would have never been accomplished. 
	
	I also received invaluable feedback throughout graduate school from members of the Farley Lab. Their conversations, encouragement, and camaraderie throughout the years helped me get through some of the toughest moments of my life. Thanks to \textit{Benjamin Song} for being the best experimental scientist and person I have ever been blessed to cross paths with—and for generating all of the data for our joint thesis projects. Our projects had highs and (many) lows, but we made it to the end! I would also like to thank \textit{Granton Jindal}, \textit{Genevieve Ryan}, \textit{Hannah Finnegan}, \textit{Jesse “Joe” Solvason}, \textit{Alexis Bantle}, \textit{Jessica Grudzien}, \textit{Sophia Le}, \textit{Joanna Encarnacion}, \textit{Krissie Tellez}, and \textit{Fabian Lim}. Also, a special shout out to members of the Evo/Devo Journal Club-–including \textit{Professors Jim Posakony} and \textit{Michael Perry}, as well as their students–-for teaching me how to properly synthesize research papers, even if it meant learning more about insects than I ever would have expected on a medical school campus.
	
	Next, I would like to thank \textit{Professor Hannah Carter}, \textit{Michelle Dow}, Andrea Castro, \textit{Kivilcim Ozturk}, \textit{Meghana Pagadala}, \textit{Clarence Mah}, \textit{James Talwar}, \textit{Adam Klie}, \textit{Cameron Waller}, \textit{Sunduk Hwang}, \textit{Jeanna Sheen}, and \textit{David Laub} for welcoming me as an honorary member of their lab. There were many days during graduate school when I looked forward to coming to the (mostly windowless) huts on the medical school campus just so I could chat with them about anything on my mind. From debating the benefits of specific neural network models to growing my wine palette over holiday dinners to drinking copious amounts of sparkling water, you all made my graduate school experience much more enjoyable!  
	
	During my second year of graduate school, I had the opportunity to join the Genetics Training Program (GTP). Even though I had to roll out of bed at times for the early morning—8:00 AM each Wednesday—Genetics Journal Club, I am thankful to \textit{Professor Bruce Hamilton} for teaching the other GTP students and me how to think like a better researcher over free bagels (but no coffee, sadly). I’ll cherish our journal club's fruitful discussions about statistics, data visualization, experimental design, and scientific ethics for years to come.
	
	Student advocacy and support is something I will always hold dear to my heart, and thus, I would also like to recognize the individuals I worked with on various program-specific and general UCSD campus initiatives. Despite not being overtly related to research, the student experience is important, and I worked closely with the Graduate Bioinformatics Council (GBIC) to help build a better community for my current and future peers. Thank you to \textit{Daniela “Dana” Nachmanson}, \textit{Jonathan Pekar}, \textit{Clarence Mah}, \textit{Jennifer Havens}, \textit{Owen Chapman}, \textit{Hannah Mummey}, \textit{Caitlin Guccione}, and many others for spending time with me writing emails, sitting in meetings, and planning events for our students with limited budgets. I would also like to thank the multitude of people I worked with during my years in the Graduate and Professional Student Association (GPSA) as a member of the Finance Committee, as Finance Committee Chair, and finally as Vice President of Financial Affairs. Thank you to everyone in GPSA for supporting me and countless other individuals across the UCSD campus, especially \textit{Graduate Divison Dean James “Jim” Antony}, \textit{John Hughes}, \textit{Breana Clark}, \textit{Anna Dickson}, \textit{Krish Bhutwala}, \textit{Andy Ryan}, \textit{Ximena Garcia-Arceo}, \textit{Hema Kopalle}, \textit{Mia Rose}, \textit{Chiaki Santiago}, \textit{Hayden Schill}, \textit{Linda Li}, \textit{Angus Chapman}, \textit{Kristin Leadbetter}, \textit{Sushil S}, \textit{Ross Turner}, \textit{Ben Du}, \textit{Giulia Corno}, \textit{Matthew Fain}, \textit{Joseph Rainaldi}, \textit{Becca Rose}, \textit{Kane Wu}, and \textit{Gabriel Zalles-Ballivian}. 
	
	Before graduate school, I was fortunate to work with the knowledgeable people at Agena Bioscience at their San Diego headquarters. Without the encouragement of people at Agena, specifically in Assays by Agena (AbA) and in Molecular Tools, I would have never considered applying to graduate school in the first place, nor would I have gained the confidence I have today as a bioinformatician. I would, however, like to give a special thanks to \textit{Julie Vanhnasy} and \textit{Huimin “Helen” Tao}. Julie and Helen were my first two managers coming out of college, and I wouldn’t have had it any other way. They were the most encouraging people I’ve had the pleasure of working with, and under their wings, I learned how to be a confident scientist for the first time.
	
	I have been blessed to have been surrounded by friends throughout graduate school from all stages of my life. Thank you to my friends from back home in the Bay Area for always being there for me, even though we are barely able to meet up online or in person due to the craziness of adult life, especially \textit{Luna Chang}, \textit{Rebecca Tien}, \textit{Justin "JP" Petrola}, and members of the Sisterhood of the Traveling Yoga Pants. Thank you to my friends from my time at UCSD who call me crazy for still being here—yes, I still, unfortunately, eat at Price Center-—including \textit{Karina Kak}, \textit{Jenny Woo}, \textit{Tiffany Diep}, \textit{Victoria Nguyen}, \textit{Larry Zhang}, \textit{Victor Wong}, \textit{Richard Phouasalith}, and \textit{Kritin Karkare}. Thank you to my friends from the Llama Lair and Litmas Lair for always giving me something exciting to do while stressing out about graduate school in our old Berwick Drive home. Thank you to the Internet crew for supporting me when IRL was too hard, including the Early 2000s Tumblr Crew, the Danger Rangers, the Cute Keyboard Club, and many others—you know who you are, even if I don’t mention you by name or online handle. Thank you to my fellow Bioinformatics \& Systems Biology peers for making graduate school lively, especially \textit{Jonathan Pekar}, \textit{Gibraan Rahman}, \textit{Adam Officer}, \textit{George Armstrong}, \textit{Jennifer Havens}, \textit{Adam Jussila}, \textit{Clarence Mah}, \textit{Cameron Martino}, \textit{Anthony Aylward}, \textit{Owen Chapman}, \textit{Carlos Guzman}, \textit{Kiki Spaulding}, \textit{Jessica Au}, \textit{Avery Pong}, \textit{Alex Jambor}, \textit{Xiaomi Du}, \textit{Emily Kobayashi}, and the entire “BISB et al., 2018” cohort-or at least whose names I didn't already mention here. I would also like to thank \textit{Ashley Tess} from the Chemistry Doctoral Program, as well as \textit{Natalie Deforest}, \textit{Sara Elmsaouri}, \textit{Jenna Kovsky}, \textit{Danielle Schafer}, \textit{James Yu}, and other students from the 2018 Cohort of the Biomedical Sciences program that adopted me during late nights at Mesa Rim or during hangouts on Beeramar. 
	
	I would like to thank my best friend and loving partner, \textit{Clarence Mah}, for being my rock throughout my entire time at UCSD—from 2013 to today. Clarence has unconditionally encouraged me to pursue my passions and been there to see me grow, and I will forever be grateful to have somebody willing to listen to my bad jokes and help me take care of Yuuki when she’s being an especially sassy Shiba Inu. 
	
	Finally, I would like to thank my wonderful but enormous family for supporting my studies throughout the years, even though they sometimes found the whole process confusing and convoluted. I would especially like to thank my parents, \textit{Francisco} and \textit{Lea Ragsac}, and my younger brother, \textit{Thomas “JonJon” Ragsac}, for the millions of phone calls and food. I am also grateful for the support of my \textit{Lola}, \textit{Josefina “Fina” Reyes}, and late \textit{Lolo}, \textit{Hermogenes “Moneng” Reyes}, to pursue anything that makes me happy. There are also the cousins, \textit{titos}, and \textit{titas} whose names I cannot mention here because that document would end up being a thesis in and of itself. 
	
	This time of my life was one of the most difficult intellectually, physically, and mentally, and it is not without the support of an entire village that I could be where I am today. There are probably many names that I neglected to mention, but even if you’re not listed, \textbf{\textit{maraming salamat}}.
	
	Chapter 1, in full, has been submitted for publication of the material as it may appear in “Diverse logics encode notochord enhancers.” Benjamin P. Song, Michelle F. Ragsac, Krissie Tellez, Granton A. Jindal, Jessica L. Grudzien, Sophia H. Le, Emma K. Farley. \textit{Cell Reports}, 2022. The dissertation author was the primary investigator and co-first author of this paper.
\end{acknowledgements}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Vita
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{vita}
\noindent
\begin{cv}{}
\begin{cvlist}{}
	\item[2013--2017] B.S. in Bioengineering: Bioinformatics\\
		University of California San Diego
	\item[2015--2016] Writing Studio Mentor, Sixth College Writing Studio\\
		University of California San Diego
	\item[2017] Teaching Assistant\\Scripps Institute of Oceanography 
	\item[2017] Research Associate, Applications and Technology Services--Assays by Agena\\
		Agena Bioscience
	\item[2018] Research Associate, Research and Development--Molecular Tools\\
		Agena Bioscience 
	\item[2018--2022] Ph.D. in Bioinformatics \& Systems Biology,\\
		University of California San Diego
	\item[2018] Teaching Assistant\\
		Scripps Institute of Oceanography
	\item[2020--2021] Teaching Assistant, School of Medicine\\
		University of California San Diego 
	\item[2020--2021] Bootcamp Instructor, Bioinformatics \& Systems Biology Doctoral Program\\
		University of California San Diego 
\end{cvlist}
\end{cv}

%%%%% PUBLICATIONS %%%%%
\publications{}

\noindent \textit{Author names marked with $\dagger$ indicate shared first co-authorship.} \newline
\noindent \textit{Publications marked with $\triangle$ are included in this text.} \newline

\noindent Granton A. Jindal$\dagger$, Alexis T. Bantle$\dagger$, Joe J. Solvason$\dagger$, Jessica L. Grudzien, Agnieszka D'Antonio-Chronowska, Fabian Lim, Sophia H. Le, \textbf{Michelle F. Ragsac}, Benjamin P. Song, Reid O. Larsen, Adam Klie, Kelly A. Frazer, and Emma K. Farley. ``Affinity-optimizing variants within cardiac enhancers disrupt heart development.'' \textit{In Submission}, 2022. \newline

\noindent Benjamin P. Song$\dagger$, \textbf{Michelle F. Ragsac}$\dagger$, Krissie Tellez, Granton A. Jindal, Jessica L. Grudzien, Sophia H. Le, and Emma K. Farley. ``Diverse logics encode notochord enhancers.'' \textit{Accepted in Principle at Cell Reports}, 2022. $\triangle$ \newline

\noindent Sydney C. Morgan$\dagger$, Stefan Aigner$\dagger$, Catelyn Anderson$\dagger$, Pedro Belda-Ferre$\dagger$, Peter De Hoff$\dagger$, Clarisse A. Marotz$\dagger$, Shashank Sathe$\dagger$, Mark Zeller$\dagger$, Noorsher Ahmed, Xaver Audhya, Nathan A. Baer, Tom Barber, Bethany Barrick, Lakshmi Batachari, Maryann Betty, Steven M. Blue, Brent Brainard, Tyler Buckley, Jamie Case, Anelizze Castro-Martinez, Marisol Chacón, Willi Cheung, LaVonnye Chong, Nicole G. Coufal, Evelyn S. Crescini, Scott DeGrand, David P. Dimmock, J. Joelle Donofrio-Odmann, Emily R. Eisner, Mehrbod Estaki, Lizbeth Franco Vargas, Michele Freddock, Robert M Gallant, Andrea Galmozzi, Nina J. Gao, Sheldon Gilmer, Edyta M. Grzelak, Abbas Hakim, Jonathan Hart, Charlotte Hobbs, Greg Humphrey, Nadja Ilkenhans, Marni Jacobs, Christopher A. Kahn, Bhavika K. Kapadia, Matthew Kim, Sunil Kurian, Alma L. Lastrella, Elijah S. Lawrence, Kari Lee, Qishan Liang, Hanna Liliom, Valentina Lo Sardo, Robert Logan, Michal Machnicki, Celestine G. Magallanes, Clarence K. Mah, Denise Malacki, Ryan J. Marina, Christopher Marsh, Natasha K. Martin, Nathaniel L. Matteson, Daniel J. Maunder, Kyle McBride, Bryan McDonald, Daniel McDonald, Michelle McGraw, Audra R. Meadows, Michelle Meyer, Amber L. Morey, Jasmine R. Mueller, Toan T. Ngo, Julie Nguyen, Viet Nguyen, Laura J. Nicholson, Alhakam Nouri, Victoria Nudell, Eugenio Nunez, Kyle O’Neill, R. Tyler Ostrander, Priyadarshini Pantham, Samuel S. Park, David Picone, Ashley Plascencia, Isaraphorn Pratumchai, Michael Quigley, \textbf{Michelle Franc Ragsac}, Andrew C. Richardson, Refugio Robles-Sikisaka, Christopher A. Ruiz, Justin Ryan, Lisa Sacco, Sharada Saraf, Phoebe Seaver, Leigh Sewall, Elizabeth W. Smoot, Kathleen M. Sweeney, Chandana Tekkatte, Rebecca Tsai, Holly Valentine, Shawn Walsh, August Williams, Min Yi Wu, Bing Xia, Brian Yee, Jason Z. Zhang, Kristian G. Andersen, Lauge Farnaes, Rob Knight, Gene W. Yeo, Louise C. Laurent. ``Automated, miniaturized, and scalable screening of healthcare workers, first responders, and students for SARS-CoV-2 in San Diego County.'' \textit{In Submission,} 2022.

%%%%% FIELDS OF STUDY %%%%% 
% This section was omitted within this thesis as it was optional.

\end{vita}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Dissertation Abstract 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{dissertationabstract}
	Embryonic development across all vertebrates begins upon the fertilization of an egg by a sperm cell to become a single-celled zygote. Embryogenesis continues with various stages of division to eventually make up an entire organism. The processes governing development are finely orchestrated and include many participants, such as genes involved in gene regulatory networks and non-coding regions of DNA, or enhancers, to regulate the expression of those genes. Defects or perturbations to this strictly regulated machinery can lead to various clinical conditions, such as congenital heart disease. Thus, deepening our understanding of embryogenesis may help us understand the mechanisms driving congenital abnormalities as well as the evolution of developmental pathways. One defining characteristic of all chordate embryos is the presence of a notochord during development. The notochord is a long, semi-rigid fibrous rod of mesodermal origin that provides structural support and serves as a signaling center to pattern the neighboring neural tube, paraxial mesoderm, and gut. A complete understanding of notochord structure and function during early and late life stages is thus essential to better understand congenital vertebral defects. For example, failure of vertebral notochord cells to transition to the nucleus pulposus, the cushioning between intervertebral discs of the spine, is associated with chordomas, slow-growing tumors formed from notochord cell remnants within the spine or the base of the skull. The ascidian \textit{Ciona intestinalis Type A} (\textit{Ciona}) is a marine organism that is evolutionarily similar to vertebrates. Through electroporation, \textit{Ciona} is readily amenable to high-throughput, high-resolution functional studies of cis-regulatory elements like enhancers in their native, whole-embryo context. To identify key notochord enhancers, I analyzed the importance of enhancer grammar–the transcription factor order, orientation, spacing, and binding affinity–in modulating notochord-specific expression. Next, I highlight the potential of single-cell RNA-sequencing to study the gene regulatory networks governing notogenesis and their relationship to congenital abnormalities. This body of work provides new insight into the regulatory processes governing notochord development, providing direction for future efforts to improve our understanding of notochord-based diseases across chordates. Finally, I highlight Open Educational Resources (OERs) I developed for Bioinformatics education, emphasizing accessibility and inclusion.
\end{dissertationabstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Main Text of the Dissertation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\mainmatter{}

\include{1_introduction}
\include{2_diverse-logics-encode-notochord-enhancers}
\include{3_proof-of-concept-grammar-searches}
\include{4_single-cell-ciona-gastrulation}
\include{5_bioinformatics-education}
\include{6_epilogue}

% Command that provides dummy text to showcase formatting:
% \Blinddocument

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Appendix of the Dissertation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

\include{2_diverse-logics-encode-notochord-enhancers_appendix}
\include{3_proof-of-concept-grammar-searches_appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% End of the Dissertation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\backmatter{}
\bibliographystyle{unsrtnat} 
\bibliography{bibliography}

\end{document}
